Stage 1 Registered Report: The error when inferring phylogenies with incipient species by a Birth-Death model

### 2018-03-29 article.tex
Wrote introduction

% Affiliations
\usepackage{authblk}
\title{The error when inferring phylogenies with incipient species by a Birth-Death model}
% \subtitle{Should protracted speciation be incorporated in phylogenetic tree construction methods?}

@@

\begin{abstract}

  The construction of phylogenies helps us answer evolutionary biological
  questions.
  Our current phylogenetic tools ignore the fact that speciation takes time,
  which has an effect unknown in phylogeny reconstruction.
  Here, we simulate true incipient species trees and their corresponding
  DNA alignments, from which we measure the reconstruction of the phylogeny
  using a standard birth-death model.
  We measure the errors that the Bayesian phylogenetic software tool BEAST2
  gives when recovering simulated phylogenies, for different times-to-speciate,
  under a range of additional parameter settings.
  It has been found that branch lengths are consistently and strongly 
  underestimated for biologically relevant parameters.
  This research shows that protractedness is a complexity of nature that
  cannot always be ignored and should be incorporated in our phylogenetic
  tools.

\end{abstract}

{\bf Keywords:} computational biology, evolution, phylogenetics, BEAST2, R

@@

What is the effect of sampling either the youngest or oldest species-to-be
to represent a species?

What is the effect of protractedness on the inference error we make?

>>>>

% Affiliations
\usepackage{authblk}
\title{Stage 1 Registered Report: The error when inferring phylogenies with incipient species by a Birth-Death model}
% \subtitle{Should protracted speciation be incorporated in phylogenetic tree construction methods?}

@@

@@

Phylogenetics enables us to infer which species lived when and
estimate how long it took for a species to give rise to a new
lineage. Phylogenies show us taxa that have completed speciation, yet
those taxa only: species that are in the process of speciation are 
overlooked.

There are multiple reasons why a speciation event, during
or after its course, goes unnoticed. It may be that the build-up
of reproductive isolation is still in an intermediate stage. But it
may also be that speciation has completed, but we, humans, have not
yet observed it. One example are the four African giraffe species, 
that have been thought to have been one single species up until 2016,
where the MRCA of the two most recent species goes back 1.6 million years.

It took some time until the Birth-Death model, a phylogenetic model
assuming constant rates of speciation and extinction, got extended
by a version that incorporates the fact that speciation takes time.
This extension is called the Protracted Birth-Death model. In this
model, a branching event does not give rise to a new species, but to
a new species-to-be, called an incipient species. Such an incipient
species may go extinct, or finish its speciation to become a good species.

Although we know speciation takes time, most of our theoretical 
phylogenetic modeling ignores this fact. To the best of our knowledge,
there is no tool incorporating this fact. One reason may be to ease
computation, enabling researchers to work on bigger phylogenies (with
more power statistical power) instead. Additionally, speciation being
protracted is only playing a role in the present. As seen from the present,
all older species have finished speciation, as we -by definition- recognize
these species as such.

There are problems with ignoring protracted speciation. A very pragmatic
example is the conservation of the now four smaller groups of giraffes,
instead of one big group. But from a more theoretical perspective, the
observed decline of speciation rate in time (excluding the contemporary human-driven
mass extinction event) may be explained by these new species not being 
recognized yet. Another problem is that the MRCA of young species is estimated 
incorrectly, as in the BD model a species can be ultra-young (as speciation
is assumed to be instantaneous), where the PBD model will place these MRCAs
further back in time (as two events need to have finished completion).

From a theoretical point of view, there is another reason why protracted
speciation has yet become a member of our toolbox: there is no framework
where it fits in: a framework assumes either an analysis at the species or
subspecies level. When picking a tool that does its analysis at the
species level, there is no protocol how to select which incipient species
gets to represent the full species it is still a member of. When picking
a tool at the sub-species level, the incipient state disappears as a
sub-species lineage, and the process degrades to an ordinary birth-death 
process.

This research's goals are two-fold, which are (1) to quantify the
inference error we make with contemporary tools assuming a Birth-Death
process, and (2) to establish a protocol how to sample incipient species
to represent a species in such a tool.

To quantify the inference error, we simulate 'true' incipient species
trees, sample a 'true' species trees from those, simulate a DNA alignments 
that follows a species tree's ontogeny, then infer a posterior of
phylogenies from each alignment. The 'true' species tree is compared to
the posterior's phylogenies with the nLTT statistic, which can be viewed
as a measure of similarity. We expect this error to increase with an
increasing protractednedness (that is, a low speciation completion rate),
but as the extent of this error has never been measured, we will be the
first to measure it.

To establish a protocol how to sample incipient species to represent species,
we compare two sampling methods that are opposites of one another: one
method is to select the oldest sub-species member to represent a species,
the other method picks the youngest sub-species member. As we do not have
a way to determine which sampling method will be better, all we determine
is if the sampling methods are different. We expect these
sampling methods to be equivalent. This research measures the extent to
which this is true.
